 Casting multiple shadows: interactive data visualisation with tours and embeddings

### 2020-02-06 paper.Rmd
write up t-SNE section

As part of a visualization workflow, it's important to consider the perception 
and interpretation of embedding methods as well. @Sedlmair2013-pn showed that 2D
scatter plots were mostly suffucient for detecting class seperation, however
they noted that often multiple embeddings were required. For the task of cluster
identification, @Lewis2012-ai showed experimentally that novice
users of non-linear embedding techniques were more likely to consider clusters
of points on a 2d scatter plot to be the result of a spurious embedding
compared to advanced users who were aware of the inner workings of the embedding
algorithm. @Wattenberg2016-ji noted that in the case of t-distributed stochastic
neighbourhood embeddings (t-SNE) it was important for users to modify algorithm
parameters in order to preserve global structure in the low dimensional
embedding.

Embedding methods can be classified into two broad groups: linear
and non-linear methods. Linear methods perform a linear transformation
of the data; one example is principal components analysis (PCA)
which performs an eigendecomposition of the estimated sample covariance matrix.
The eigenvalues are sorted in decreasing order and represent the variance
explained by each component (eigenvector). A common approach to deciding on
the number of principal components to retain is to plot the proportion of variance explained by each component and choose a cut-off. Non-linear methods
generally perform preprocessing on the high-dimensional data such as generating
a neighborhood graph and perform transformations on the preprocessed form. We
restrict our attention to three methods that are commonly used in
high-throughput biology (though this is by no means exhaustive): t-SNE, 
uniform manifold alignment and projection (UMAP), and PHATE.

Clearly, finding an appropriate embedding is a difficult and somewhat
poorly defined problem. We propose a more pragmatic workflow inspired by (wickham blindfold paper) by incorporating
interactive graphics and tours with embeddings that allows users to see a
global overview of their high dimensional data and assists them with cluster
orientation tasks. This workflow is incorporated into an R package
called `liminal` (Available: https://github.com/sa-lee/liminal).

>>>>

As part of a visualization workflow, it's important to consider the perception 
and interpretation of embedding methods as well. @Sedlmair2013-pn showed that 2D
scatter plots were mostly suffucient for detecting class seperation, however
they noted that often multiple embeddings were required. For the task of cluster
identification, @Lewis2012-ai showed experimentally that novice
users of non-linear embedding techniques were more likely to consider clusters
of points on a 2d scatter plot to be the result of a spurious embedding
compared to advanced users who were aware of the inner workings of the embedding
algorithm.

Embedding methods can be classified into two broad groups: linear
and non-linear methods. Linear methods perform a linear transformation
of the data; one example is principal components analysis (PCA)
which performs an eigendecomposition of the estimated sample covariance matrix.
The eigenvalues are sorted in decreasing order and represent the variance
explained by each component (eigenvector). A common approach to deciding on
the number of principal components to retain is to plot the proportion of variance explained by each component and choose a cut-off.

Non-linear methods
generally perform preprocessing on the high-dimensional data such as generating
a neighborhood graph and perform transformations on the preprocessed form. We
restrict our attention to three methods that are commonly used in
high-throughput biology: t-distributed stochastic neighbor embedding (t-SNE), 
uniform manifold alignment and projection (UMAP), and PHATE. The t-SNE algorithm estimates the similarity of (Euclidean) distances of points in a high dimensional space using a Gaussian distribution and then  
estimates a configuration in the low dimensional embedding space by modelling
similarities using a t-distribution with 1 degree of freedom. The resulting
configuration is the one that minimizes the Kulback-Leibler divergence
between the two distributions. A recent theoretical contribution by @Linderman2019-dq proved that t-SNE can recover spherical and well separated 
cluster shapes, and proposed new approaches for tuning the optimisation parameters. It is a known problem that t-SNE can have trouble recovering
global structure and that configurations can be highly dependent on
how the algorithm is initialised and parameterised 
[@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
related to LargeVis, and like t-SNE acts on the k-nearest neighbor graph. Its
main differences are that it uses a different cost function (cross entropy) which is optimized using stochastic gradient descent and defines a different
kernel for similiarites in the low dimensional space. Finally, PHATE is based

There is no one-size fits all: finding an appropriate embedding for a given
dataset is a difficult and somewhat poorly defined problem. For non-linear methods, there are a lot of parameters to explore that can have an effeft on the resulting visualisation and interpretation. We propose a more pragmatic workflow inspired by (wickham blindfold paper) by incorporating
interactive graphics and tours with embeddings that allows users to see a
global overview of their high dimensional data and assists them with cluster
orientation tasks. This workflow is incorporated into an R package
called `liminal` (Available: https://github.com/sa-lee/liminal).

https://jlmelville.github.io/smallvis/init.html#other_initialization_options
